Data cleaning Study 4

### 2024-09-02 appendix_exp1.Rmd
draft before edits Hugo

@@

>>>>

## Participants

We recruited 200 participants from the US via prolific. 6 participants failed our attention check, resulting in a final sample of `r exp1_descriptives$n_subj` participants (`r exp1_descriptives$gender$female$n` female, `r exp1_descriptives$gender$male$n` male; $age_\text{mean}$: `r exp1_descriptives$age$mean`, $age_\text{sd}$: `r exp1_descriptives$age$sd`, $age_\text{median}$: `r exp1_descriptives$age$median`). Since we did not have any prior assumptions on effect sizes, we did not do a power analysis.

@@

## Results

Regarding RQ1 and RQ2, participants answered on average `r exp1_descriptives$means$knowledge_mean` (sd = `r exp1_descriptives$means$knowledge_sd`) of the questions correctly, and accepted the scientific consensus on average for `r exp1_descriptives$means$acceptance_mean` (sd = `r exp1_descriptives$means$acceptance_sd`) of the questions.

Fig. \@ref(fig:exp1-conditional-acceptance) illustrates the relationship between knowledge and acceptance. In most cases (`r exp1_descriptives$conditional_acceptance$false$Yes$share`), participants readily accepted the scientific consensus after having given the wrong answer to a question. In very few cases (`r exp1_descriptives$conditional_acceptance$true$No$share`), participants who gave the correct response afterwards rejected the scientific consensus, thereby contradicting their own initial response. We believe this might have been due to inattention.

For RQ3, we find a positive but small correlation between both science knowledge and trust in science (r = `r exp1_descriptives$cor_trust_knowledge$estimate`, p `r exp1_descriptives$cor_trust_knowledge$p.value`), and acceptance of scientific consensus and trust in science (r = `r exp1_descriptives$cor_trust_acceptance$estimate`, p `r exp1_descriptives$cor_trust_acceptance$p.value`). The more people are knowledgeable about science and the more they tend to accept the scientific consensus, the more they tend to trust science. These correlations are relatively weak, which might be partly due to ceiling effects: As illustrated in Fig. \@ref(fig:exp1-plot), (i) most people do trust science, and (ii) that is true even among people with low knowledge or acceptance rates.

For RQ4, we find a negative correlation of similar magnitude between conspiracy thinking and science knowledge (r = `r exp1_descriptives$cor_conspiracy_knowledge$estimate`, p `r exp1_descriptives$cor_conspiracy_knowledge$p.value`), and conspiracy thinking and acceptance of scientific consensus (r = `r exp1_descriptives$cor_conspiracy_acceptance$estimate`, p `r exp1_descriptives$cor_conspiracy_acceptance$p.value`).

In Appendix \@ref(exp1), we show that these associations are robust when using alternative measures of trust and conspiracy thinking. Appendix \@ref(exp1) also includes more descriptive statistics, such as knowledge and acceptance by science questions.

Are trust in science and conspiracy thinking, respectively, associated with being more easily convinced of the scientific consensus? In our main analyses, we looked at correlations of acceptance across all observations. One possibility is that the associations between trust in science/conspiracy thinking and acceptance of scientific consensus are explained by science knowledge: People who give the right answer in the first place are more ready to accept the consensus, and trust in science/conspiracy thinking are mostly associated with this knowledge, but not with willingness to accept the consensus. To addressed this potential confound, in a non-preregistered analysis, we restricted our sample to cases where participants gave the wrong answer to the knowledge question. We then calculated the correlation between trust in science and average acceptance rate by participant. We find no statistically significant correlation of acceptance with neither conspiracy thinking (r = `r exp1_descriptives$false_answers_cor_conspiracy_acceptance$estimate`, p `r exp1_descriptives$false_answers_cor_conspiracy_acceptance$p.value`), nor with trust in science (r = `r exp1_descriptives$false_answers_cor_trust_acceptance$estimate`, p `r exp1_descriptives$false_answers_cor_trust_acceptance$p.value`).

## Discussion

These results suggest that most people accept the scientific consensus most of the time. Even when people do not know the correct answer to a science question, they tend to mostly accept the scientific consensus afterwards. Yet, in `r exp1_descriptives$conditional_acceptance$false$No$share` of the cases, participants rejected the scientific consensus after having given the wrong answer, suggesting that simply stating the consensus is not sufficient to convince participants sometimes. In general, people with lower trust in science and who believe more in conspiracy theories tend to both know less about science and accept the scientific consensus less.

### 2024-09-02 appendix_exp2.Rmd
draft before edits Hugo

@@

@@

>>>>

In experiment 1, we tested whether participants would accept the scientific consensus on basic science facts. In most instances they did, but not always. In experiment 2, we wanted to test whether this reluctance was because of participants not trusting us as a source of consensual science knowledge. To do so, we added an explanation and sources to each consensus statement, instead of only stating the consensus. To better understand reasons for consensus rejection, after having answered all questions, we asked participants an open-ended question to explain why they rejected the consensus, for each question on which they did so. We also excluded the where do trees their materials from, as this question clearly seemed to be an outlier where most participants would get the answer wrong (see Appendix \@ref(exp1)).

Based on experiment 1, we formulated the following hypotheses:

**H1a: Higher trust in science is associated with more science knowledge?**

**H1b: Higher trust in science is associated with more acceptance of the scientific consensus, for participants who did not already know it?**

**H2a: Higher conspiracy thinking is associated with less science knowledge?**

**H2b: Higher conspiracy thinking is associated with less acceptance of the scientific consensus, for participants who did not already know it?**

We had the following research questions:

**RQ1: What is the average science knowledge score?**

**RQ2: When a participant’s answer does not match the consensus, how often do they change their mind and accept the consensus?**

**RQ3: What reasons do participants provide to justify their rejection of the scientific consensus?**

## Participants

We recruited 201 participants from the US via prolific. 11 participants failed our attention check, resulting in a final sample of `r exp2_descriptives$n_subj` participants (`r exp2_descriptives$gender$female$n` female, `r exp2_descriptives$gender$male$n` male; $age_\text{mean}$: `r exp2_descriptives$age$mean`, $age_\text{sd}$: `r exp2_descriptives$age$sd`, $age_\text{median}$: `r exp2_descriptives$age$median`). Since we did not have any prior assumptions on effect sizes and our analyses were descriptive, we did not do a power analysis.

## Procedure

The procedure was the same as in experiment 1, with the difference that, instead of just stating the scientific consensus, participants were presented with a short explanation which we wrote, partly based on explanations generated by ChatGPT, and three links to authoritative sources supporting the answer (Fig. \@ref(fig:exp2-stimulus-example).

@@

We relied on the same items as in experiment 1. The only difference was that we removed the question on trees.

@@

## Results

As in experiment 1, we find a positive but small correlation between science knowledge and trust in science (H1a: r = `r exp2_descriptives$cor_trust_knowledge$estimate`, p `r exp2_descriptives$cor_trust_knowledge$p.value`) and a small negative correlation between science knowledge and conspiracy thinking (H2a: r = `r exp2_descriptives$cor_conspiracy_knowledge$estimate`, p `r exp2_descriptives$cor_conspiracy_knowledge$p.value`). By contrast to experiment 1, we conditioned on initially false answers when looking at the relationship of consensus acceptance with trust in science and conspiracy thinking, respectively. For trust in science, we find no statistically significant correlation (r = `r exp2_descriptives$false_answers_cor_trust_acceptance$estimate`, p `r exp2_descriptives$false_answers_cor_trust_acceptance$p.value`). For conspiracy thinking we find a small negative one (r = `r exp2_descriptives$false_answers_cor_conspiracy_acceptance$estimate`, p `r exp2_descriptives$false_answers_cor_conspiracy_acceptance$p.value`).

Confirming results from experiment 1, we find that the more people are knowledgeable about science and the more they tend to accept the scientific consensus even when they are not that knowledgeable in science, the more they tend to trust science. These correlations are relatively weak, which might be partly due to ceiling effects: As illustrated in Fig. \@ref(fig:exp2-plot), (i) most people do trust science, and (ii) that is true even among people with low knowledge or acceptance rates. In Appendix \@ref(exp2) we show that these results hold for our alternative measures of trust and conspiracy thinking. We also include more descriptive statistics, such as knowledge and acceptance by science questions.

Regarding RQ1, participants answered on average `r exp2_descriptives$means$knowledge_mean` (sd = `r exp2_descriptives$means$knowledge_sd`) of the questions correctly, and accepted the scientific consensus on average for `r exp2_descriptives$means$acceptance_mean` (sd = `r exp2_descriptives$means$acceptance_sd`) of the questions. Fig. \@ref(fig:exp2-conditional-acceptance) illustrates the relationship between knowledge and acceptance. In response to RQ2, in most cases (`r exp2_descriptives$conditional_acceptance$false$Yes$share`), participants readily accepted the scientific consensus after having initially given the wrong answer to a question. In very few cases (`r exp2_descriptives$conditional_acceptance$true$No$share`), participants who gave the correct response afterwards rejected the scientific consensus, thereby contradicting their own initial response.

```{r exp2-justifications}
exp2_justifications %>%
  group_by(category)%>%
  summarize(n = n(), 
            n_subjects = n_distinct(id)) %>%
  mutate(Share = n/sum(n)) %>%
  mutate_if(is.numeric, round, digits=3) %>% 
  mutate(Share = paste0(Share*100, "%")) %>% 
  arrange(desc(n)) %>% 
  select(category, n, Share, n_subjects) %>% 
  rename(Category = category,
         `N (instances)` = n, 
         `Share (instances)` = Share, 
         `N (unique participants)` = n_subjects
         ) %>% 
  apa_table(caption = "Justifications by category")
```

For RQ3, we got `r exp2_descriptives$justifications_n` answers from `r exp2_descriptives$justifications_n_participants` different participants to the open-ended questions on why they had rejected the scientific consensus on a particular question. Table \@ref(tab:exp2-justifications) summarizes these answers by five categories. All answers can be read in Appendix \@ref(exp2).

## Discussion

Similar to experiment 1, most people (i) do know and agree with the scientific consensus, and (ii) tend to accept the scientific consent even if they were not previously aware of it (i.e. answered the knowledge question wrongly). The share of these latter is considerably larger in experiment 2 (`r exp2_descriptives$conditional_acceptance$false$Yes$share`) than in experiment 1 (`r exp1_descriptives$conditional_acceptance$false$Yes$share`). While this could be just sampling variation, it might be that adding explanations and sources convinced people more than merely stating the consensus. We also show, again, that people with lower trust in science and who believe more in conspiracy theories tend to both know less about science and accept the scientific consensus less.

### 2024-09-02 exp_4/cleaning.Rmd
draft before edits Hugo

## Import data
```{r import, message=FALSE}
d <- read_csv("data/test.csv")
names(d)
```

@@

There are % failed attention checks.

@@

```{r}

d_long <- d_wide %>% 
  # bring to long format
  pivot_longer(cols = c(
    ends_with("know"),
    ends_with("accept"),
    ends_with("confirm"),
    ends_with("link1"),
    ends_with("link2")
  ), 
  names_to = "subject_question", values_to = "answer") %>% 
  # separate subject (e.g. water) and question (e.g. acceptance)
  separate_wider_delim(subject_question, "_", names = c("subject", "question")
                       ) %>%
  pivot_wider(names_from = question, values_from = answer) %>% 
  # make a click count variable of the sum of both link clicks
  mutate(
    # make link variables numeric
    across(contains("link"), ~as.numeric(.x)),
    # make sum variable
    link_clicks = link1 + link2) %>% 
# create better variable names
  rename(knowledge = know, 
         acceptance_initial = accept, 
         acceptance_confirmation = confirm) %>% 
  # make a final acceptance variable
  mutate(acceptance = ifelse(is.na(acceptance_confirmation), acceptance_initial, acceptance_confirmation))
```

@@

>>>>

## Import data
```{r import, message=FALSE}
d <- read_csv("data/qualtrics.csv")
names(d)
```

@@

There are 2 failed attention checks.

@@

```{r}
d_long <- d_wide %>% 
  # bring to long format
  pivot_longer(cols = c(
    ends_with("know"),
    ends_with("accept"),
    ends_with("confirm"),
    ends_with("link1"),
    ends_with("link2")
  ), 
  names_to = "subject_question", values_to = "answer") %>% 
  # separate subject (e.g. water) and question (e.g. acceptance)
  separate_wider_delim(subject_question, "_", names = c("subject", "question")
                       ) %>%
  pivot_wider(names_from = question, values_from = answer) %>% 
  # make a click count variable of the sum of both link clicks
  mutate(
    # make link variables numeric
    across(contains("link"), ~as.numeric(.x)),
    # make sum variable
    link_clicks = link1 + link2) %>% 
# create better variable names
  rename(knowledge = know, 
         acceptance_initial = accept, 
         acceptance_confirmation = confirm) %>% 
  # make a final acceptance variable
  mutate(acceptance = ifelse(is.na(acceptance_confirmation), acceptance_initial, acceptance_confirmation))
```

@@

```{r}
# Extract justifications
# questions <- unique(data_long$subject)
# 
# justifications <- d_wide %>%
#   select(id, all_of(questions)) %>%
#   pivot_longer(all_of(questions),
#                names_to = "question",
#                values_to = "answer") %>%
#   drop_na(answer) %>%
#   arrange(id)
# 
# write_csv(justifications, "data/justifications.csv")
```
